---
title: "PracticalMachineLearningProject"
author: "Dxander"
date: "Saturday, January 24, 2015"
output: pdf_document
---

Executive Summary

The purpose of this project is to analyze the activity of 6 participants to determine how well they perform activities.  The data are from accelerometers on the belt, forearm, and dumbell use of the 6 participants.  They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.  

Information on other uses of the data for this project can be found at http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

This project involves classification.  The goal is to classify the the manner in which the participant did the exercise.  
If the exercise was performed exactly to specifications (Class A), throwing the elbows to the front (Class B), 
lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and 
throwing the hips to the front (Class E).  See (VBGUF 2015)


Key points in the process are 1) Classify the manner in which the exercise was done.  2) The classification model should be built using features and cross-validataion.  3) As a performance measure calculate the out of sample error.  4) Finally, use the test data and the model to correctly classify 20 test cases.

A resource for this project is the paper by Velloso, E; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitive Activity Recognition of Weight Lifting Exercises (2013)


# Data Setup
Load required packages and set random number generator to ensure reproduceability.

```{r setlibrary}
library(Hmisc)
library(caret)
library(kernlab)
library(randomForest)
library(corrplot)
set.seed(9237)
```

set the working directory

```{r setwd}
setwd("E:/DataScientist/PracticalMachineLearning/Project")
```

```{r downloaddata}
downloadDataset <- function(URL="", destFile="data.csv"){
        if(!file.exists(destFile)){
                download.file(URL, destFile, method="curl")
        }else{
                message("You already downloaded the data!")
        }
}

trainURL<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

testURL <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
downloadDataset(trainURL, "pml-training.csv")

downloadDataset(testURL, "pml-testing.csv")
```

# Load the data into R
The original data set consist of lots of data marked "NA" and summary data that is not
in the testing set.   This data is removed from the training set.
Same procedures are performed on both training and final testing set.
```{r intoR}
training <- read.csv("pml-training.csv", na.strings=c("#DIV/0!","NA", "") )
final_testing <- read.csv("pml-testing.csv", na.strings=c("#DIV/0!", "NA", "") )
dim(training)
dim(final_testing)
summary(training)
```

```{r exploreData}
training <- training[ ,colSums(is.na(training)) == 0]
final_testing <- final_testing[ ,colSums(is.na(testing)) == 0]

for(i in c(8:ncol(training)-1)) {training[,i] = as.numeric(as.character(training[,i]))}
for(i in c(8:ncol(final_testing)-1)) {final_testing[,i] = as.numeric(as.character(testing[,i]))}

training   <- training[,-c(1:7)]
final_testing <- final_testing[,-c(1:7)]
```


Examine the feature set and create the model data
```{r featureset}
feature_set <- colnames(training[colSums(is.na(training)) == 0])[-(1:7)]
model_data <- training[feature_set]
feature_set

inTrain <- createDataPartition(y=model_data$classe, p=0.6, list=FALSE )
training <- model_data[inTrain,]
testing <- model_data[-inTrain,]
dim(training); dim(testing);
```

```{r plot}
summary(training$classe)
plot(training$classe, col=rgb((1:5)/5,0,(1:5)/5), main="Frequency Plot of Classe variable")
```

# Correlated Predictors
```{r correlations}
M <- abs(cor(training[,-length(training)]))
diag(M) <- 0
which(M > 0.8, arr.ind=T)

correlMatrix <- cor(training[, -length(training)])
corrplot(correlMatrix, order = "FPC", method = "circle", type = "lower", tl.cex = 0.8,  tl.col = rgb(0, 0, 0))

```
Although there are many correlated predictors because I am not a specialist in sports physics I do not know of 
a way to combine the variables in a knowledgeable way.   So I will continue with the main tasks.


# Construct model with 4-fold cross validation
```{r buildmodel}
model <- train(training$classe ~., data= training, method="rf", prox=TRUE,
               trControl = trainControl(method = "cv", number =4, allowParallel=TRUE))

model
```
# To calculate the prediction accuracy of the classification model
# 1st classify training set
```{r predict1}
train_pred <- predict(model, training)
confusionMatrix(train_pred, training$classe)
```
For the training set the in sample accuracy is 1 or 100%


# 2nd classify testing set or out of sample accuracy
```{r predict2}
test_pred <- predict(model, testing)
confusionMatrix(test_pred, testing$classe)
```
For the test set the out of sample accuracy is 0.998 or 99.8%



## Prediction Assignment
```{r predictFinal}
answers <- predict(model, final_testing)
answers <- as.character(answers)
answers

# Using the test program code provided
pml_write_files = function(x){
        n = length(x)
        for(i in 1:n){
                filename = paste0("problem_id_",i,".txt")
                write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
        }
}



pml_write_files(answers)
```

